Probing the internal structures of quasars using microlensing events

### 2016-03-14 paper1/intro.tex
edits

Active galactic nucleii are thought to be powered through the
accretion of matter from the proximal environment into a supermassive
black hole.  The radiation emitted excites the surrounding medium
which becomes detectable as narrow line regions, broad line regions
and optical continuum.  Moreover, in the direction perpendicular to
the accretion disk, where the medium is more transparent, jets will
appear.  If a jet is oriented towards the Earth, a quasar is observed
\citep[e.g.,][]{1984RvMP...56..255B}.  While the basic mechanism
\citep[originating in the work
  of][]{1964ApJ...140..796S,1964SPhD....9..246Z,1969Natur.223..690L}
is not in doubt, the central engines, near the event horizons of the
black holes, remain to be probed.

For the black holes in the Galactic centre and the centre of M87
---the two nearest objects that barely qualify as AGN--- the central
engines (which are $<0.1\rm\,mas$ on the sky) are close to being
resolved through very long baseline interferometry, which shows
preliminary indications of the jet-launching structures
\citep{2008JPhCS.131a2055D,2012Sci...338..355D,2013MNRAS.434..765K,2016arXiv160205527F}.
Current data do not deliver images, but require fitting to to
predefined models for the images.  A whole range of models have been
applied, starting from simple geometric models to more complex
physical models
\citep{2008Natur.455...78D,2011ApJ...738...38B,2009ApJ...706..497M,2010ApJ...717.1092D}.
The more complicated models nonetheless tend to predict a crescent
shaped silhouette of the black hole.  This motivated
\cite{2013MNRAS.434..765K} to use a simple geometric crescent model to
fit the data, and argued that the crescent is nothing but the
silhouette of the event horizon.

@@

Gravitational microlensing of quasars, reviewed in
Section~\ref{sec:microlensing} below, refers to sharp changes in the
observed brightness of quasars that have been lensed by intervening
galaxies, without any changes in the intrinsic luminosity.
Microlensing affects only the light from the innermost part of the
quasar, such as the optical continuum
\citep[e.g.,][]{2012A&A...544A..62S} and is a consequence of two
things: the very small size of the central engine, and granularity of
the mass distribution of a lensing galaxy due to stars.  The latter
means that the brightness of a lensed object (or lensing
magnification) is not a smooth function of source position, but a
contains a complicated network of singular curves, known as caustics.
Figure~\ref{fig:magnification_map} shows part of a magnification map
with a few caustics.  The lensed brightness would be given by placing
the source on such a magnification map and integrating the surface
brightness weighted by the magnification.  Most astrophysical sources
straddle several caustics, and hence their net brightness varies
smoothly with location.  The central engine of quasars, however, is
smaller than the typical spacing between caustics.  As a result, the
lensed brightness undergoes sudden changes as a quasar crosses a
caustic.  The effect supplies an upper limit on the size of the
central engine, and can also be used to study the mass distribution
and kinematics of stars in the lensing galaxy as well
\citep[e.g.,][]{2012ApJ...744..111P}.  Caustics have an additional
remarkable feature: though they can be very complicated, they have
some universal properties well-known from catastrophe theory.  In
particular, very close to the simplest caustics (known as folds), the
magnification is approximately constant on one side and
$\propto1/\sqrt p$ where $p$ is the transverse distance of the source
from the caustic.  This property will be exploited later.

@@

The crescent characteristics identified in
Figure~\ref{fig:char_points} are from an idealised source crossing an
idealised caustic.  In Section~\ref{sec:numerics} we see what happens
when we replace both of these with numerical simulations.  The ideal
fold-transit is replaced by the path AB in
Figure~\ref{fig:magnification_map}.  The simple crescent is replaced
by a source (Figure~\ref{fig:M87_image}) from published simulations of
the black-hole environment of M87
\citep{2009MNRAS.394L.126M,2012MNRAS.421.1517D}.  The simulated light
curve so obtained (lower panel of Figure~\ref{fig:M87_plots}) still
shows, albeit faintly, features from the crescent.

@@

>>>>

Active galactic nuclei are thought to be powered through the
accretion of matter from the proximal environment into a supermassive
black hole.  The radiation emitted excites the surrounding medium
which becomes detectable as narrow line regions, broad line regions
and optical continuum.  Moreover, in the direction perpendicular to
the accretion disk, where the medium is more transparent, jets will
appear.  If a jet is oriented towards the Earth, a quasar is observed
\citep[e.g.,][]{1984RvMP...56..255B}.  While the basic mechanism
\citep[originating in the work
  of][]{1964ApJ...140..796S,1964SPhD....9..246Z,1969Natur.223..690L}
is not in doubt, the central engines, near the event horizons of the
black holes, remain to be probed.

For the black holes in the Galactic centre and the centre of M87
---the two nearest objects that barely qualify as AGN--- the central
engines (which are $<0.1\rm\,mas$ on the sky) are close to being
resolved through very long baseline interferometry, which shows
preliminary indications of the jet-launching structures
\citep{2008JPhCS.131a2055D,2012Sci...338..355D,2013MNRAS.434..765K,2016arXiv160205527F}.
Current data do not deliver images, but require fitting to
predefined models for the images.  A whole range of models have been
applied, starting from simple geometric models to more complex
physical models
\citep{2008Natur.455...78D,2011ApJ...738...38B,2009ApJ...706..497M,2010ApJ...717.1092D}.
The more complicated models nonetheless tend to predict a crescent
shaped silhouette of the black hole.  This motivated
\cite{2013MNRAS.434..765K} to use a simple geometric crescent model to
fit the data, and argued that the crescent is nothing but the
silhouette of the event horizon.

@@

Gravitational microlensing of quasars, reviewed in
Section~\ref{sec:microlensing} below, refers to sharp changes in the
observed brightness of quasars that have been lensed by intervening
galaxies, without any changes in the intrinsic luminosity.
Microlensing affects only the light from the innermost part of the
quasar, such as the optical continuum
\citep[e.g.,][]{2012A&A...544A..62S} and is a consequence of two
things: the very small size of the central engine, and granularity of
the mass distribution of a lensing galaxy due to stars.  The latter
means that the local magnification is not a smooth function of source
position. It contains a complicated network of singular curves, known
as caustics.  Figure~\ref{fig:magnification_map} shows part of a
magnification map with a few caustics.  The lensed brightness would be
given by placing the source on such a magnification map and
integrating the surface brightness weighted by the magnification.
Most astrophysical sources straddle several caustics, and hence their
net brightness varies smoothly with location.  The central engine of
quasars, however, is smaller than the typical spacing between
caustics.  As a result, the lensed brightness undergoes sudden changes
as a quasar crosses a caustic.  The effect supplies an upper limit on
the size of the central engine, and can also be used to study the mass
distribution and kinematics of stars in the lensing galaxy as well
\citep[e.g.,][]{2012ApJ...744..111P}.  Caustics have an additional
remarkable feature: though they can be very complicated, they have
some universal properties well-known from catastrophe theory.  In
particular, very close to the simplest caustics (known as folds), the
magnification is approximately constant on one side and
$\propto1/\sqrt p$ where $p$ is the transverse distance of the source
from the caustic.  This property will be exploited later.

@@

The simple crescent is replaced
by a source (Figure~\ref{fig:M87_image}) from published simulations of
the black-hole environment of M87
\citep{2009MNRAS.394L.126M,2012MNRAS.421.1517D}.  The simulated light
curve so obtained (lower panel of Figure~\ref{fig:M87_plots}) still
shows, albeit faintly, features from the crescent.
The crescent characteristics identified in
Figure~\ref{fig:char_points} are from an idealised source crossing an
idealised caustic.

In Section~\ref{sec:numerics} we see what happens
when we replace both of these with numerical simulations.  The ideal
fold-transit is replaced by the path AB in
Figure~\ref{fig:magnification_map}.

@@

Finally in Section~\ref{sec:discussion} we discuss in more depth the
implications of the results presented in the previously mentioned
sections.  One of the most interesting implication is the possibility
to estimate the black hole mass from the reconstructed
parameters. This further requires good approximations of the relative
transversal velocity between the quasar and the lens. Proper
constraints can be set with independent observations of the stellar
structure that contains the gravitational lens.

The thorough study of the possibility to reconstruct the quasar's
structural parameters from light-curves containing multiple
microlensing events represents the target of future work. This will
most probably require the use of powerful statistical tools. Another
path for future work is the designing of an observation regime best
suited for acquiring the necessary microlensed light-curve.

### 2016-03-14 paper1/ms.tex
another set of corrections
Small title typo
edits

\title{Microlensing as a possible probe of the internal structure of quasars}

\author[Tomozeiu et al]{Mihai Tomozeiu$^{1,2}$,Irshad Mohammed$^1$, Manuel Rabold$^{1,2}$,
\newauthor
Prasenjit Saha$^1$ and Joachim Wambsganss$^3$\\
$^1${Physik-Institut, University of Zurich, Winterthurerstrasse 190,
  8057 Zurich, Switzerland} \\
$^2${Institute for Computational Science, University of Zurich,
  Winterthurerstrasse 190, 8057 Zurich, Switzerland} \\
$^3${Zentrum f\"ur Astronomie der Universit\"at Heidelberg,
  M\"onchhofstrasse 12--14, 69120 Heidelberg, Germany}
}

@@

\input{code}

\input{fitting}

>>>>

%\title[Quasar microlensing and its internal structure]{Properties of quasar microlensing light curves: probing its internal structure}
\title{Probing the internal structures of quasars using microlensing events}

\author[Tomozeiu et al]{Mihai Tomozeiu\thanks{mihai@physik.uzh.ch},$^{1,2}$ 
Irshad Mohammed,$^{1,2,3}$ 
Manuel Rabold,$^{1,2}$
Prasenjit Saha,$^{1,2}$  
\newauthor
and Joachim Wambsganss$^4$\\
$^1${Physik-Institut, University of Zurich, Winterthurerstrasse 190,
  8057 Zurich, Switzerland} \\
$^2${Institute for Computational Science, University of Zurich,
  Winterthurerstrasse 190, 8057 Zurich, Switzerland} \\
$^3${Theoretical Astrophysics Group, Fermi National Accelerator Laboratory, Batavia, IL 60510, USA}\\
$^4${Zentrum f\"ur Astronomie der Universit\"at Heidelberg,
  M\"onchhofstrasse 12--14, 69120 Heidelberg, Germany}
}

@@

\input{numerics}

### 2016-03-14 paper1/numerics.tex
edits

\section{Model fitting and likelihood analysis}\label{sec:fitting}

In this section we simulate the fitting of source models to
lightcurves in the presence of random and systematic errors.  We
generate mock data for three source models --- a crescent, a uniform
disc and a Gaussian disc --- and then fit each mock light curve to all
three source models. Parameter fitting and marginalising was done by
Markov-chain Monte Carlo (MCMC).

@@

\subsection{Mock datasets}

@@

\subsection{Fitting to mock data sets}

>>>>

\section{Fitting mock data}\label{sec:numerics}

Having used the simple model of an ideal fold to gain insight, we now
move to the numerical computation of microlensing light curves with
realistic mass distributions, and then fit source models to
lightcurves in the presence of random and systematic errors.

First we generate microlensing magnification maps using the
microlensing code by \cite{1999A&A...346L...5W}.  We then generate
mock light curves for three source models --- a crescent, a uniform
disc and a Gaussian disc --- and then fit each mock light curve to all
three source models. Parameter fitting and marginalising was done by
Markov-chain Monte Carlo (MCMC).

\subsection{Numerical microlensing magnification maps}

The code uses a ray-shooting technique to compute the gravitational
lensing effect of a mass distribution consisting of (a)~a smooth
component and (b)~a random distribution of point masses representing
stars.  The ray shooting maps a grid of $\vec\theta$ to $\vec\beta$
using the lens equation (\ref{eqn:lens}).  That is, the rays are shot
from the observer back to the source.

For the computation of the individual deflection angles
$\vec\alpha(\vec\theta)$ a heirarchical or tree method is used.  The
positions of all lensing masses are put into a grid of $\vec\theta$.
Each grid cell is subdivided into four smaller squares recursively
until every cell contains only one mass.  Nearby masses are added
individually, while distance masses are clumped into larger grid cells
whose net contribution is approximated by its first few multipole
moments.  Scaled units are used, with the constant pre-factor in the
deflection angle (\ref{eqn:alpha}) separated out.

The result of ray shooting is a pixel map on the $\vec\beta$ plane of
the number of lightrays which arrive at the source plane from a
particular observer.  This intermediate result is effectively a
magnification map on the source plane.  Once the map is created, the
lightcurve can be obtained by specifying the transit path of the
source across the map.  At each point on this transit line, the code
computes equation (\ref{eqn:ft2d}) convolving the brightness
distribution of the source with the magnification pattern of the map.
In real life, not only are both lens and source moving, but the lens
lens configuration, and with it the magnification pattern, is also
changing with time.  While the first subtlety is taken care of by a
coordinate transformation in this analysis, for the second one the
lens configuration is assumed to be constant in time.

When generating the magnification map depicted in
Figure~\ref{fig:magnification_map}, which is used in our analysis
below, only two point masses were included.  This was done in order to
have clean fold caustics.  For the computation of the actual
lightcurve, the code was modified to also allow for crescent shaped
images specified through the parameter set $R_p,R_n,a,b$. Here $R_p$
denotes the outer radius of the crescent and $R_n$ the inner one. The
orientation of the source image with respect to the magnification map
is specified by the parameters $a$ and $b$ as the shift of the center
of the inner disk from the center of the outer disk in $x$- and
$y$-direction respectively. In the original version of the code,
gaussian and disk shaped images where already implemented. Those are
completely characterised by the single parameter $R_p$. The values of
the parameters are specified in pixel units corresponding to the
magnification map. Further one needs to specify the start and end
point coordinates of the path, which the center of the source image
follows through the magnification map (see the depiction in figure
\ref{fig:magnification_map}). Hence the points along this path are
specified through the number of timesteps for which the computation of
the brightness is to be carried out. Those points correspond to the
actual measurement of the brightness of an object in the observational
case. For each timestep the two-dimensional convolution of equation
\ref{eqn:ft2d} is carried out numerically for the position of the
source on the magnification map.

For the purpose of this analysis it was desirable to mimic the
analytical behaviour of a simple fold as much as possible, for
comparing the numerical result with the analytical one, therefore the
path of the source was chosen, so that it intersects the border of the
caustic perpendicularly, and on a point where the border is a fold
caustic.

@@

\subsection{Mock light curves}

@@

\subsection{Model fitting and likelihood analysis}
